Reanalysis Step 5: Outcome Analysis with Three-Way Interaction

### 2025-09-01 reanalysis-5-outcome_analysis-interaction.qmd
bug fix

@@

## Average Marginal Effects by Subgroup

```{r ame-calculation}
# Function to calculate and pool AMEs across imputations
calculate_pooled_ames <- function(imputed_datasets, models, n_imputations) {

  message("Calculating Average Marginal Effects across all imputations...")

  # Store AME results from each imputation
  all_ame_results <- list()

  for (imp in 1:n_imputations) {
    message("  Processing imputation ", imp, "/", n_imputations)

    data_for_ame <- imputed_datasets[[imp]]
    model_for_ame <- models[[imp]]

    imp_ame_results <- data.frame()

    for (dr_level in c("No_DR", "NPDR", "PDR")) {
      for (tx_status in c("No_Treatment", "Any_Treatment")) {

        # Get subset for this subgroup
        subset_data <- data_for_ame[data_for_ame$dr_severity == dr_level & 
                                    data_for_ame$any_treatment == tx_status, ]
        n_subset <- nrow(subset_data)

        if (n_subset > 0) {
          # Create counterfactual datasets
          subset_data_0 <- subset_data
          subset_data_0$ever_lapse_binary <- factor("No_Lapse", 
                                                    levels = c("No_Lapse", "Lapsed"))

          subset_data_1 <- subset_data
          subset_data_1$ever_lapse_binary <- factor("Lapsed", 
                                                    levels = c("No_Lapse", "Lapsed"))

          # Get predictions
          pred_0 <- predict(model_for_ame, newdata = subset_data_0, type = "response")
          pred_1 <- predict(model_for_ame, newdata = subset_data_1, type = "response")

          # Calculate individual effects
          individual_effects <- as.numeric(pred_1) - as.numeric(pred_0)

          # Calculate weighted AME
          weights <- subset_data$iptw_weight
          ame_value <- weighted.mean(individual_effects, w = weights, na.rm = TRUE)

          # Weighted variance for SE
          weighted_var <- sum(weights * (individual_effects - ame_value)^2) / sum(weights)
          se_value <- sqrt(weighted_var / n_subset)

          # Store results
          imp_ame_results <- rbind(imp_ame_results, data.frame(
            DR_Severity = dr_level,
            Treatment = tx_status,
            N = n_subset,
            AME = ame_value,
            SE = se_value,
            stringsAsFactors = FALSE
          ))
        }
      }
    }

    all_ame_results[[imp]] <- imp_ame_results
  }

  # Pool AME results across imputations
  # Get unique subgroups
  first_df <- all_ame_results[[1]]
  n_subgroups <- nrow(first_df)

  pooled_ame <- data.frame(
    DR_Severity = first_df$DR_Severity,
    Treatment = first_df$Treatment,
    N = first_df$N,  # Should be same across imputations
    AME = numeric(n_subgroups),
    SE = numeric(n_subgroups),
    stringsAsFactors = FALSE
  )

  # Apply Rubin's rules to each subgroup
  for (j in 1:n_subgroups) {
    # Extract AMEs and SEs for this subgroup
    ames <- sapply(all_ame_results, function(x) x$AME[j])
    ses <- sapply(all_ame_results, function(x) x$SE[j])

    # Rubin's rules
    Q_bar <- mean(ames)
    U_bar <- mean(ses^2)
    B <- var(ames)
    T <- U_bar + (1 + 1/n_imputations) * B

    pooled_ame$AME[j] <- Q_bar
    pooled_ame$SE[j] <- sqrt(T)
  }

  # Add confidence intervals and p-values
  pooled_ame <- pooled_ame |>
    mutate(
      CI_Lower = AME - 1.96 * SE,
      CI_Upper = AME + 1.96 * SE,
      P_Value = 2 * pnorm(-abs(AME / SE))
    )

  return(pooled_ame)
}

# Calculate pooled AMEs
ame_results <- calculate_pooled_ames(imputed_datasets, models, n_imputations)

# Format for display
ame_display <- ame_results |>
  mutate(
    AME = round(AME, 4),
    SE = round(SE, 4),
    CI = paste0("(", round(CI_Lower, 4), ", ", round(CI_Upper, 4), ")"),
    P_Value = format.pval(2 * pnorm(-abs(AME/SE)), digits = 4, eps = 0.0001)
  ) |>
  dplyr::select(DR_Severity, Treatment, N, AME, SE, CI, P_Value)

kable(ame_display,
      caption = "Average Marginal Effects of Lapsing by Subgroup",
      col.names = c("DR Severity", "Treatment", "N", "AME", "SE", "95% CI", "P-value"),
      row.names = FALSE)
```

### Forest Plot of Average Marginal Effects

```{r ame-forest-plot}
#| fig-height: 6
#| fig-width: 10

# Prepare data for forest plot
forest_data <- ame_results |>
  mutate(
    # Create a combined label for the y-axis
    Subgroup = paste0(DR_Severity, " + ", Treatment),
    # Order for plotting (reverse for forest plot convention)
    Subgroup = factor(Subgroup, levels = rev(unique(Subgroup)))
  ) |>
  arrange(desc(Subgroup))

# Create forest plot
forest_plot <- ggplot(forest_data, aes(x = AME, y = Subgroup)) +
  # Add reference line at 0 (no effect)
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray50", alpha = 0.7) +

  # Add confidence intervals
  geom_errorbarh(aes(xmin = CI_Lower, xmax = CI_Upper), 
                 height = 0.2, size = 0.8, color = "gray30") +

  # Add point estimates
  geom_point(aes(color = DR_Severity, shape = Treatment), size = 4) +

  # Labels and theme
  labs(
    title = "Average Marginal Effects of Lapsing on Vision Impairment",
    subtitle = "By DR severity and treatment status (with 95% CI)",
    x = "Average Marginal Effect (Probability Scale)",
    y = "",
    color = "DR Severity",
    shape = "Treatment"
  ) +

  # Theme adjustments
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 12),
    axis.text.y = element_text(size = 11, hjust = 0),
    axis.text.x = element_text(size = 11),
    axis.title = element_text(size = 12),
    legend.position = "bottom",
    legend.box = "horizontal",
    panel.grid.major.y = element_line(color = "gray90"),
    panel.grid.minor = element_blank()
  ) +

  # Color scale
  scale_color_manual(values = c(
    "No_DR" = "#1f77b4",
    "NPDR" = "#ff7f0e", 
    "PDR" = "#d62728"
  )) +

  # Shape scale
  scale_shape_manual(values = c(
    "No_Treatment" = 16,  # Circle
    "Any_Treatment" = 17   # Triangle
  )) +

  # X-axis formatting
  scale_x_continuous(labels = scales::percent_format(accuracy = 0.1))

# Add text annotations with exact values
forest_plot_annotated <- forest_plot +
  geom_text(aes(label = sprintf("%.3f", AME)),
            nudge_y = 0.3, size = 3.5, color = "black")

# Display the annotated plot
forest_plot_annotated
```

@@

```{r export}
# Save key results for reporting
results_to_save <- list(
  pooled_coefficients = pooled_results,
  subgroup_effects = subgroup_effects_stored,
  ame_results = ame_results,
  probability_table = if(exists("prob_table")) prob_table else NULL,
  model_summaries = model_summaries,
  n_imputations = n_imputations,
  analysis_mode = analysis_mode,
  timestamp = Sys.time()
)

>>>>

# CACHE SETTINGS - Set these to TRUE to force recalculation
force_refit_models <- FALSE  # Set to TRUE to re-fit all models
force_recalc_emmeans <- FALSE  # Set to TRUE to recalculate marginal means

@@

@@

```{r export}
# Note: Using emmeans-based risk differences as the primary effect measure
# These are calculated at the reference grid rather than averaged over observed data

# Save key results for reporting
results_to_save <- list(
  pooled_coefficients = pooled_results,
  subgroup_effects = subgroup_effects_stored,
  emmeans_results = pooled_results,  # The emmeans-based effects
  probability_table = if(exists("prob_table")) prob_table else NULL,
  model_summaries = model_summaries,
  n_imputations = n_imputations,
  analysis_mode = analysis_mode,
  timestamp = Sys.time()
)
